---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysing the ONS 'GVA by industry by ITL region' data in R

The ONS produces an [Excel document](https://www.ons.gov.uk/economy/grossvalueaddedgva/datasets/nominalandrealregionalgrossvalueaddedbalancedbyindustry) with GVA data at three different geographical scales, ITL1 to ITL3. Wikipedia has an [excellent explanation with maps](https://en.wikipedia.org/wiki/International_Territorial_Level) of the ITL regions.

Here, we'll explore some ways to analyse this data using R. The data has been processed to make it more useable in R - see the [process_from_excel.R]([process_from_excel.R) script and its comments for an explanation of how to get from the Excel sheet to the datasets used here. All of the derived files (and the original ONS excel sheet, currently used version dated 25th April 2023) are included in this repository in the [data folder](https://github.com/DanOlner/regionalGVAbyindustry/tree/master/data). Test relative data folder link: [data folder](data). 


```{r iris}
summary(iris)

plot(iris$Sepal.Length, iris$Sepal.Width)
```